# Data

## Technical Description

The SLD dataset (url: https://edg.epa.gov/EPADataCommons/public/OA/EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv) was gathered by EPA (United States Environmental Protection Agency), which summarizes more than 90 different indicators associated with the built environment and location efficiency.Most attributes are available for all U.S. block groups.

EPA first released a beta version of the Smart Location Database in 2011. The initial full version was released in 2013, and the database was updated to its current version in 2021.

The format of the data is geometry data based on census block groups. We plan to select the features that can best describe our research question, and select the data falls into Manhatten area in New York city. We plan to plot the density distribution of the data to examine the usability of the features for future use.

### Data dimension
The original SLD dataset dimension is Rows: 220740 Columns: 117, each row stands for each census block group unit. The columns include 90 indicators and other geometry IDs.

```{r}

library(readr)
library(DataExplorer)
library(ggplot2)
library(naniar)
library(VIM) 

url <- "https://edg.epa.gov/EPADataCommons/public/OA/EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv"


sld2021 <- read_csv(url)





```

### Feature selection
We choose 32 features out of 90 to best describe the research question, all the features are from five prospects:demographics, density, design, transit access, destination accessibility.

Demographics
‘P_WrkAge’, Percent of population that is working aged 18 to 64 years, 
Pct_AO1 ,Percent of one-car households in CBG, 2018 
Pct_AO2p ,Percent of two-plus-car households in CBG, 2018 
R_HiWageWk, Count of workers earning $3333/month or more (home location), 2017
E_HiWageWk, # of workers earning $3333/month or more (work location), 2017


Density
D1a, Gross residential density (HU/acre) on unprotected land 
D1b, Gross population density (people/acre) on unprotected land
D1c, Gross employment density (jobs/acre) on unprotected land 
D1c8_Hlth , Gross health care (8-tier) employment density (jobs/acre) on unprotected land
D1d , Gross activity density (employment + HUs) on unprotected land 
D2a_JpHH, Jobs per household 
D2a_EpHHm,  Employment and household entropy 
D2c_TripEq, Trip productions and trip attractions equilibrium index; the closer to one, the more balanced the trip making
D2r_JobPop, Regional Diversity. Standard calculation based on population and total employment: Deviation of CBG ratio of jobs/pop from the regional average ratio of jobs/pop 
D2c_WrEmIx, Household Workers per Job Equilibrium Index; the closer to one the more balanced the resident workers and jobs in the CBG. 

Design
D3a, Total road network density 
D3aao, Network density in terms of facility miles of auto-oriented links per square mile
D3amm, Network density in terms of facility miles of multi-modal links per square mile 
D3apo, Network density in terms of facility miles of pedestrian- oriented links per square mile

D3b ,Street intersection density (weighted, auto-oriented intersections eliminated) 
D3bao ,Intersection density in terms of auto-oriented intersections per square mile
D3bmm4, Intersection density in terms of multi-modal intersections having four or more legs per square mile
D3bpo4, Intersection density in terms of pedestrian-oriented intersections having four or more legs per square mile 

Transit Access 
D4a, Distance from the population-weighted centroid to nearest transit stop (meters) ,2020
D4c, Aggregate frequency of transit service within 0.25 miles of CBG boundary per hour during evening peak period, 2020
D4d, Aggregate frequency of transit service [D4c] per square mile 

Destination Accessibility 
D5ar, Jobs within 45 minutes auto travel time, time- decay (network travel time) weighted, 2020
D5ae, Working age population within 45 minutes auto travel time, time-decay (network travel time) weighted 
D5br , Jobs within 45-minute transit commute, distance decay (walk network travel time, GTFS schedules) weighted 
D5be,  Working age population within 45-minute transit commute, time decay (walk network travel time, GTFS schedules) weighted 
D5cr, Proportional Accessibility to Regional Destinations - Auto: Employment accessibility expressed as a ratio of total CBSA accessibility
D5cri, Regional Centrality Index – Auto: CBG [D5cr] score relative to max CBSA [D5cr] score 

```{r}


selected_columns <- c("GEOID20", "STATEFP", "COUNTYFP", "TRACTCE", "BLKGRPCE", 
                      "P_WrkAge", "Pct_AO1", "Pct_AO2p", "R_HiWageWk", 
                      "E_HiWageWk", "D1A", "D1B", "D1C", "D1C8_HLTH", "D1D", 
                      "D2A_JPHH", "D2A_EPHHM", "D2C_TRIPEQ", "D2R_JOBPOP", 
                      "D2C_WREMLX", "D3A", "D3AAO", "D3AMM", "D3APO", "D3B", 
                      "D3BAO", "D3BMM4", "D3BPO4", "D4A", "D4C", "D4D", 
                      "D5AR", "D5AE", "D5BR", "D5BE", "D5CR", "D5CRI")

new_sld2021 <- sld2021[, selected_columns]

```
### Area selection
We filtered the Manhatten data by selecting data with "STATEFP"==36, and "COUNTYFP"==61.The "STATEFP"==36 indicates the New York state, the "COUNTYFP"==61 is Manhatten area.
```{r}

NY_data <- new_sld2021[new_sld2021$STATEFP == 36, ]

man_data <- NY_data[NY_data$COUNTYFP == 61, ]

```


## Missing value analysis
```{r}


selected_columns <- c("P_WrkAge", "Pct_AO1", "Pct_AO2p", "R_HiWageWk", 
                      "E_HiWageWk", "D1A", "D1B", "D1C", "D1C8_HLTH", "D1D", 
                      "D2A_JPHH", "D2A_EPHHM", "D2C_TRIPEQ", "D2R_JOBPOP", 
                      "D2C_WREMLX", "D3A", "D3AAO", "D3AMM", "D3APO", "D3B", 
                      "D3BAO", "D3BMM4", "D3BPO4", "D4A", "D4C", "D4D", 
                      "D5AR", "D5AE", "D5BR", "D5BE", "D5CR", "D5CRI")
man_miss<- man_data[, selected_columns]
missing_plot <- plot_missing(man_miss,
  group = list(Good = 0.05, OK = 0.4, Bad = 0.8, Remove = 1),
  group_color = list(Good = "#1B9E77", OK = "#E6AB02", Bad = "#D95F02", Remove =
    "#E41A1C"),
  missing_only = FALSE, title = "Distribution of Missing Values for Each Variable")

```
There are no missing values in the dataset, as all features have 0 missing rows. In the plot_missing() function, when no missing values are present, the graph defaults to an x-axis scale ranging from -0.05 to 0.05. A similar case is demonstrated in the official reference:  https://cran.r-project.org/web/packages/DataExplorer/vignettes/dataexplorer-intro.html

```{r}
#install.packages("naniar")

vis_miss(man_miss)

```
We also use heatmap for missing value by packages("naniar"), from the observations, 100% present of data are exsited. 0% percent value is missing.

```{r}



aggr(man_data, col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, 
     labels = names(man_data), cex.axis = 0.7, gap = 3, 
     main = "Missing Data Overview")
```
There is no missing data because proportion of missings is 0. If the combination count is 1 in the output of the aggr() function, it means there is only one distinct pattern of missingness across all the variables in the dataset. In other words:Every row with missing values has missingness in the exact same set of variables.No other unique missing value patterns exist in the dataset.

