# Results

---
title: "5702 project script"
output: html_document
date: "2024-12-08"
---

```{r}
# Load necessary libraries

library(ggplot2)
library(corrplot)
library(dplyr)
library(ggridges)
library(patchwork)
library(reshape2)
library(GGally)
library(tidyr)
library(plotly)
library(sf) 
library(shiny)
```

```{r}
# Load the dataset
data <- read.csv("SLD_new.csv")
data <- data %>%
  mutate(D4A = ifelse(D4A < 0, 0, D4A))
summary(data)
```

```{r}
num_variables <- c(
  "P_WrkAge", "Pct_AO1", "R_HiWageWk", 
  "D1B", 
  "D3A", "D3AMM", "D3APO", 
  "D4A", 
  "D5AR", 
  "D5CRI","D1C", "D3B"
)


selected_data <- data[num_variables]


data_long <- selected_data %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

# Histograms with free x and y scales
ggplot(data_long, aes(x = Value)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free") +  # Allow both x and y axes to vary
  labs(
    title = "Histogram with Density Line for Numerical Variables",
    x = "Value", 
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
  )


```

## Convert D1C and D3B to categorical data

D1C (Gross Employment Density):

The histogram shows a heavily right-skewed distribution with a large concentration of values near zero and only a few observations at higher values.
This uneven distribution suggests that employment density can be categorized into groups such as "low," "medium," and "high" rather than treated as a continuous variable for easier interpretation and analysis.

D3B (Street Intersection Density):

Similar to D1C, the histogram is right-skewed, with most values clustered near the lower end and a long tail of higher values.
The nature of street intersection density makes it meaningful to group the values into categories like "low," "medium," "high," and "very high" to reflect differences in urban planning and infrastructure.

Therefore, we choose to convert this two variables to categorical variables

```{r}
# Transform D1C and D3B to categorical data
data_categorical <- data %>%
  mutate(
    D1C_Cat = cut(D1C, breaks = c(-Inf, 40, 80, 120, Inf), labels = c("Low", "Medium", "High", "Very High")),
    D3B_Cat = cut(D3B, breaks = c(-Inf, 40, 80, 130, Inf), labels = c("Low", "Medium", "High", "Very High"))
  )

#write.csv(data_categorical, "data_new.csv")

#head(data_categorical)
```

## 1. Variables distirubution

### (1) histogram for numeric variables

```{r}
num_variables <- c(
  "P_WrkAge", "Pct_AO1", "R_HiWageWk", 
  "D1B", 
  "D3A", "D3AMM", "D3APO", 
  "D4A", 
  "D5AR", 
  "D5CRI"
)


selected_data <- data[num_variables]


data_long <- selected_data %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

# histograms
ggplot(data_long, aes(x = Value)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free") +
  labs(title = "Histogram with Density Line for Numerical Variables",
       x = "Value", y = "Density") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) )

```

1. D1B (Gross population density)

2. D3A (Total road network density)

3. D3AMM (Multi-modal network density) 

4. D3APO (Pedestrian-oriented network density) 

5. D4A (Distance to nearest transit stop) 

6. D5AR (Jobs within 45 minutes auto travel time)

7. D5CRI (Regional Centrality Index) 

8. P_WrkAge (Percent of working-age population)

9. Pct_AO1 (Percent of one-car households)

10. R_HiWageWk (Count of high-wage workers earning $3,333/month or more)

The histogram plot provides a visual overview of the distributions for several numerical variables. 

D1B (Gross population density) and D3A (Total road network density) show right-skewed distributions, with most areas concentrated at lower values and a few with much higher densities. D3AMM (Multi-modal network density) and D3APO (Pedestrian-oriented network density) are also right-skewed, suggesting limited infrastructure in many regions.

D4A (Distance to nearest transit stop) exhibits a sharp peak at lower values, indicating many areas are close to transit stops, while some are much farther.

D5AR (Jobs within 45 minutes auto travel time) and D5CRI (Regional Centrality Index) show near-normal distributions, suggesting these metrics are relatively evenly distributed across regions. 

P_WrkAge (Percent of working-age population) displays a bell-shaped distribution, indicating consistency in the working-age population. Pct_AO1 (Percent of one-car households) shows a right-skewed pattern, highlighting a concentration of areas with fewer one-car households. 

Finally, R_HiWageWk (Count of high-wage workers) is highly skewed, with most areas having fewer high-wage workers and only a small fraction with significantly higher counts. 


### (2). ridgeline plot for numerical variables
```{r}
# ridgeline plot 


data_long <- data_long %>%
  group_by(Variable) %>%
  mutate(Value_scaled = (Value - min(Value, na.rm = TRUE)) / 
                         (max(Value, na.rm = TRUE) - min(Value, na.rm = TRUE))) %>%
  ungroup()


medians <- data_long %>%
  group_by(Variable) %>%
  summarize(median_value = median(Value, na.rm = TRUE)) %>%
  arrange(desc(median_value))

data_long$Variable <- factor(data_long$Variable, levels = rev(medians$Variable))

data_long <- data_long %>%
  filter(Value_scaled >= 0)



medians <- data_long %>%
  group_by(Variable) %>%
  summarize(median_value = median(Value_scaled, na.rm = TRUE)) %>%
  arrange(desc(median_value))


data_long$Variable <- factor(data_long$Variable, levels = rev(medians$Variable))


ggplot(data_long, aes(x = Value_scaled, y = Variable)) +
  geom_density_ridges(
    fill = "skyblue", alpha = 0.7, scale = 0.9, rel_min_height = 0.01
  ) +
  labs(
    title = "Ridgeline Plot for Numerical Variables Ordered by Median (Scaled Values)",
    x = "Scaled Value",
    y = "Variable"
  ) +
  xlim(0, 1) +  # Ensure the x-axis starts at 0
  theme_minimal() +
  theme(
    legend.position = "none",  
    axis.text.y = element_text(size = 10),  
    plot.title = element_text(size = 12, face = "bold")  
  )



```


The ridgeline plot shows the distribution of scaled values (0–1) for various numerical variables, ordered by their median values. 

P_WrkAge (Percent of working-age population) has the highest median and displays a centered distribution, indicating most areas have a significant working-age population. 

D5CRI (Regional Centrality Index) and D5AR (Jobs within 45 minutes auto travel time) both have high medians and fairly centralized distributions, reflecting their importance in accessibility and centrality. 
D3AMM (Multi-modal network density) and Pct_AO1 (Percent of one-car households) show moderately spread distributions, indicating diversity across regions in infrastructure and household car usage. 

D1B (Gross population density), D3A (Total road network density), and D4A (Distance to nearest transit stop) have lower medians, showing skewed distributions where certain regions dominate in these metrics.

R_HiWageWk (Count of high-wage workers) and D3APO (Pedestrian-oriented network density) have the lowest medians, with highly skewed distributions highlighting their sparsity or concentration in specific areas. 

### (3). Bar chart for categorical variables

```{r}

data_cat_freq <- data_categorical |>
  count(D1C_Cat, name = "Freq_D1C")
data_cat_freq2 <- data_categorical |>
  count(D3B_Cat, name = "Freq_D3B")

p1 <- data_cat_freq |>
  group_by(D1C_Cat) |>
  summarize(Count = sum(Freq_D1C)) |>
  ungroup() |>
  ggplot(aes(D1C_Cat, Count)) +
  labs(
    title = " Bar plot for D1C(Gross employment level)"
  ) +
  geom_col(color = "grey50", fill = "lightblue") +
  theme_minimal(base_size = 10) +  
  theme(
    plot.title = element_text(size = 10, face = "bold")  
  )

p2 <- data_cat_freq2 |>
  group_by(D3B_Cat) |>
  summarize(Count = sum(Freq_D3B)) |>
  ungroup() |>
  ggplot(aes(D3B_Cat, Count)) +
  labs(
    title = "Bar plot for D3B(Street intersection density)"
  ) +
  geom_col(color = "grey50", fill = "lightblue") +
  theme_minimal(base_size = 10) +  
  theme(
    plot.title = element_text(size = 10, face = "bold")  
  )


p1 | p2

```

Both variables show the Low level have the highest frequency, indicating that a significant proportion of observations have low values for both D1C_Cat(Gross employment level) and D3B_Cat(Street intersection density).

However, the distribution of D1C_Cat and D3B_Cat show a sharp decline in frequencies for medium and high level.


### (4). jitter graph for relationship between categorical variables
```{r}
# Add jitter plot
ggplot(data_categorical, aes(x = D1C_Cat, y = D3B_Cat)) +
  geom_jitter(width = 0.2, height = 0.2, size = 2, alpha = 0.7) +
  theme_minimal() +
  labs(
    title = "Jitter Plot for categorical variable of D1C and D3B",
    x = "D1C Category: Gross employment density level",
    y = "D3B Category: Street intersection density level"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

```

This jitter plot compares D1C (Gross employment density levels) with D3B (Street intersection density levels) by categorizing each variable into four levels: Low, Medium, High, and Very High. 

The plot reveals that there is a clear clustering within each combination of categories, with Low D1C and Low D3B showing a tight grouping of points, indicating many areas fall into this category. Similarly, Very High D1C and Very High D3B also display a dense cluster, suggesting that regions with high employment density tend to align with higher street intersection density. 

However, the spread of points increases for Medium and High categories, indicating greater variability or inconsistency in these regions. This implies a potential correlation between employment density and street intersection density, especially at extreme levels, but variability exists in intermediate categories. 


### (5). Parallel coordinate plot for all numberical variables (color by categorical variables)

We randomly select 200 sample and do parallel coordinate plot because original dataset is too large

```{r}
set.seed(123)


sampled_data <- data_categorical %>%
  sample_n(200) %>%
  select(D3B_Cat, P_WrkAge, Pct_AO1, R_HiWageWk, D1B, D3A, D3AMM, D3APO, D4A, D5AR)


numeric_vars <- sampled_data %>%
  select(-D3B_Cat)  

data_normalized <- as.data.frame(scale(numeric_vars))


data_normalized <- cbind(sampled_data["D3B_Cat"], data_normalized)


data_long <- data_normalized %>%
  mutate(ID = 1:nrow(data_normalized)) %>%
  pivot_longer(
    cols = -c(ID, D3B_Cat),
    names_to = "Variable",
    values_to = "Value"
  )


ggplot(data_long, aes(x = Variable, y = Value, group = ID, color = D3B_Cat)) +
  geom_line(alpha = 0.6) +
  labs(
    title = "Parallel Coordinate Plot for Randomly Selected 200 Rows",
    x = "Numerical Variables",
    y = "Normalized Value",
    color ="D3B: Street intersection density level"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  
    axis.text.y = element_text(size = 10),                         
    axis.title = element_text(size = 12),                         
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5), 
    legend.title = element_text(size = 8),                       
    legend.text = element_text(size = 8)                          
  )



```

This parallel coordinate plot represents the relationships between multiple normalized numerical variables for 200 randomly selected rows. Each line corresponds to one observation (row), and the colors represent the levels of the categorical variable D3B_Cat (Low, Medium, High, Very High).

Lines from Very High street intersection (purple line) generally peak at these variables, suggesting that areas with very high street intersection density level (D3B) values have higher values for transit accessibility and job accessibility.

Red lines generally stay lower, indicating that areas in the "Low" street intersection density level (D3B) are closer to transit stops and have fewer jobs accessible by car.


# The goal for our project is to see how does each factors affect the D5CRI (Regional Centrality Index) in Manhattan City

To reach our goal, we will first draw correlation plot to see how does each numerical data related to D5CRI (Regional Centrality Index) 

## 2. Correlation plot with numerical variables
### (1).heat map

```{r}


cor_matrix <- cor(selected_data, use = "complete.obs")
cor_melt <- melt(cor_matrix)


ggplot(cor_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +  
  scale_fill_gradient2(
    low = "blue", high = "red", mid = "white",
    midpoint = 0, limit = c(-1, 1), space = "Lab",
    name = "Correlation"
  ) +
  labs(
    title = "Correlation Heatmap of Numerical Variables",
    x = "Variables",
    y = "Variables"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold")
  )



```

1. Strong Positive Relationships:
Areas that are highly central (D5CRI) tend to have better access to jobs within a 45-minute drive (D5AR).


2. Moderate Positive Relationships:
Areas with more working-age residents tend to have more high-wage earners.

Areas with denser road networks also tend to have better multi-modal transportation options like buses or trains.

3. Weak or Minimal Relationships:
Walkable infrastructure (D3APO) and Places closer to transit stops (D4A) shows little connection to most other factors, indicating it operates independently.

4. Negative Relationships:
Areas designed for pedestrians tend to have slightly fewer multi-modal facilities, showing a trade-off in urban planning priorities.

### (2). Scatter plot matrix

```{r,fig.width=10,fig.height=9}



selected_data <- data[num_variables]


selected_data <- selected_data %>%
  mutate(across(everything(), as.numeric))


ggpairs(
  selected_data,
  title = "Scatter Plot Matrix of Numeircal Variables and Dependant Variables ",
  lower = list(continuous = "smooth", size = 0.01, alpha=0.5),   
  upper = list(continuous = "cor"),      
  diag = list(continuous = "densityDiag") 
)+ 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From the scatter plot we can see that transit stops, and robust transport networks are the strongest correlation with regional centrality, while factors like pedestrian infrastructure or car ownership have smaller correlation with regional centrality.

### (3). Scatter plot for D5AR(Jobs within 45 minutes) and D5CRI(Regional Centrality Index) facet on D3B (Street Intersection density level)
```{r,fig.height=8, fig.width=10}



ggplot(data_categorical, aes(x = D5AR, y = D5CRI, color = D1C_Cat)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~D3B_Cat, scales = "fixed") + 
  labs(
    title = "Scatter plot of Jobs within 45 minutes VS Regional Centrality Index by Street Intersection density level",
    x = "D5AR(Jobs within 45 minutes)",
    y = "D5CRI: Regional Centrality Index",
    color = "D1C_Cat: Employment level"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text.x = element_text(hjust = 1),
    plot.title = element_text(size = 12, face = "bold")
  )


```

This scatter plot shows a strong positive correlation between jobs within 45 minutes (D5AR) and regional centrality (D5CRI) across all street intersection density levels. Higher street intersection density levels (e.g., "Very High") exhibit more compact clustering along the trend line, indicating better regional centrality and job accessibility. Lower levels of employment density (D1C) are more scattered, particularly in the "Low" street intersection density category.

The Employment density does not seem to significantly influence the observed relationship in this data.

## 3. Map plot

```{r}

str(data_categorical)
data_trans <- data_categorical

data_trans$geometry <- st_as_sfc(data_trans$geometry, crs = 4326)
data_sf <- st_as_sf(data_trans, sf_column_name = "geometry")



#class(data_sf)
# Should include "sf"

#st_geometry_type(data_sf)

```

```{r}

data_sf <- st_cast(data_sf, "MULTIPOLYGON")  

spatial_plot <- ggplot(data = data_sf) +
  geom_sf(aes(fill = D5CRI, text = paste("Regional Centrality Index:", D5CRI)), 
          color = "grey",  
          size = 0.001,      
          alpha = 0.8      
  ) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Regional Centrality Index") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Map plot for Regional Centrality Index in Manhattan", fill = "D5CRI")


interactive_plot <- ggplotly(spatial_plot, tooltip = "text")  
interactive_plot
```

Result:
In Manhattan, the most centrally located regions generally appear around midtown—near and around the core business districts, transportation hubs, and areas with a dense street network. These midtown neighborhoods, often home to major landmarks, offices, and transit stations, stand out as the most "central" parts of Manhattan according to the map.

User Interactions:
Hover your cursor over regions to see the Regional Centrality Index values in a tooltip.
Zoom in and out or pan around the map to explore different areas more closely.
Use the toolbar in the viewer to reset the view or take screenshots as needed.


```{r}

data_sf <- st_cast(data_sf, "MULTIPOLYGON")  

spatial_plot <- ggplot(data = data_sf) +
  geom_sf(aes(fill = D5AR, text = paste("Jobs within 45 minutes:", D5AR)), 
          color = "grey",  
          size = 0.001,      
          alpha = 0.8      
  ) +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Jobs within 45 minutes") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Map plot for Jobs within 45 minutes in Manhattan", fill = "D5CRI")


interactive_plot <- ggplotly(spatial_plot, tooltip = "text")  
interactive_plot
```

The central and southern parts of Manhattan appear to have higher accessibility. This may due to the presence of major employment hubs, better transit infrastructure, and denser population centers.

This pattern occurs due to the combination of factors such as transit infrastructure (subways and buses), proximity to job centers (eg: Midtown or Financial District), and urban planning that influences accessibility within a 45-minute radius.




